\documentclass[DM,authoryear,toc]{lsstdoc}
% lsstdoc documentation: https://lsst-texmf.lsst.io/lsstdoc.html

\newcommand{\yusra}[1]{{\it \textcolor{green}{[YA: #1]}}}

\input{meta}

% To add a short-form title:
% \title[Short title]{Title}
\title{Image Display WG}

\author{%
Yusra AlSayyad (chair),
Scott Daniel,
Gregory Dubois-Felsmann,
Simon Krughoff,
Lauren MacArthur,
John Swinbank
}

\setDocRef{DMTN-126}
\setDocUpstreamLocation{\url{https://github.com/lsst-dm/dmtn-126}}

\date{\vcsDate}

\setDocAbstract{%
This document describes the findings of the LSST DM Image Display Working Group.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{1}{YYYY-MM-DD}{Unreleased.}{Yusra AlSayyad}
}


% Recommendation macro copied from dmtn-085:


% Hyperref plays havoc with my crazy recommendation linking TOCs if we let it
% turn the section name into a hyperlink, so we use the page numbers instead.
\hypersetup{linktoc=page}

\makeatletter
\newcommand{\printrecs}{%
  \section{Recommendations}%
  \label{sec:recs}
  \begin{enumerate}[leftmargin=7em,label=QAWG-REC-\arabic*:]%
  \def\@noitemerr{\@latex@warning{Empty objective list}}%
  \@starttoc{rec}%
  \end{enumerate}%
}
\newcommand{\l@rec}[2]{#1}
\newenvironment{recbox}
{
  \begin{center}
  \begin{minipage}[h]{.85\linewidth}
    \begin{tcolorbox}[colback=green!5!white,colframe=green!75!black,title=\textbf{Recommendation}]
}
{
    \end{tcolorbox}
  \end{minipage}
  \end{center}
}
%
% \begin{recommendation}{rec:label}{Brief summary}
% Explanatory text, if any.
% \end{recommendation}
\NewEnviron{recommendation}[2]
 {%
  \label{#1}%
  \addcontentsline{rec}{rec}{%
    \noexpand\unexpanded{\unexpanded\expandafter{\item{#2 (\S\ref{#1})}}}%
  }%
  \begin{recbox}
  \emph{#2.} \par \BODY%
  \end{recbox}
 }%


\begin{document}

\maketitle

\section{Introduction}

The QA Working Group (QAWG) was formed in mid-calendar-2017, with a wide-ranging remit to suggest improvements to processes and tools used for quality assurance across the Data Management subsystem, following the charge defined in \citeds{LDM-622}.
The QAWG produced an extensive report (\citeds{DMTN-085}), which touched on issues in the area of image display and visualization.
However, due to uncertainty around project scope and the future of development of image display tools within DM, the QAWG did not issue effective recommendations in this area.

The Image Display Working Group (IDWG) was constituted in summer 2019 with the primary goal of rectifying this shortcoming, as well as issuing recommendations on image display for the purposes of commissioning (which fell outside the QAWG scope) and considering whether useful recommendations can be made about tool development in support of diagnostic work within the Camera Subsystem.
The detailed terms of the Working Group are provided in \citeds{LDM-702}.

\section{Approach}

The IDWG identified three ways to approach its charge.

\begin{enumerate}

  \item{
    We considered the various contexts or environments in which a scientist or developer might wish to view images, and established how well these are served by current tools.
    For example, these environments include a developer working on a standalone laptop, or a scientist using the LSST Science Platform.
    This analysis is presented in \S\ref{sec:env}.
  }

  \item{
    We surveyed groups of key stakeholders across the LSST Project, including DM pipeline developers, members of the Commissioning Team, and members of the Camera Team, to understand what they regard as the most important use cases for visualization, and how well those use cases are served by existing tools.
    This analysis is presented in \S\ref{sec:features}.
  }

  \item{
    We identified existing tools --- including those developed within DM, those produced by other subsystems within LSST, and those available in the wider community --- to assess their capabilities and relevance to the key use cases which have been identified.
    This analysis is presented in \S\ref{sec:tools}.
  }

\end{enumerate}

Finally, in \S\ref{sec:conc} we present conclusions and recommendations drawn from the above.

\section{Environments}
\label{sec:env}
\yusra{TO DO: Simon}
There are eight environments in which a DM member would view images.

Simon's table.

Simon's chart in words.

We focused on the lsst-dev and Nublado environments, which are partially implemented.

DM requested these essential features.
Features = some action users want to be able to complete or some characteristic of the image viewer.
Use cases are the commissioning,  science validation, pipelines writing tasks that prompt these actions.

Some of these features are currently possible and easy on lsst-dev and Nublado; some are not.

We distinguish between ``possible'' and ``easy.''

\section{Features and User Stories}
\label{sec:features}
\input{features.tex}

\section{Existing Tools}
\label{sec:tools}
\input{existing_tools.tex}

\section{Recommendations}
\label{sec:conc}

\yusra{TO DO: John, Simon, Lauren, Gregory, Yusra: please review and edit}
The recommendations are based on an assessment of the requested features, supporting use cases, and existing toolset evaluated.
One currently unsupported feature is large image display.
Users cross DM and commissioning (section \ref{sec:large_viz}) will need to see a full-focal plane or full-tract image in its entirety and be able to zoom in to the individual pixels.
Of the existing tools evaluated, the LSST Camera Image Viewer was the most impressive in terms of responsiveness in the zooming.
The Camera Team reports that Tony Johnson is actively developing the viewer to include additional features.
Therefore adopting this will prevent duplication of effort.

However, the Camera Team reports that the viewer owes its responsiveness dedicated hardware.
Therefore, we recommend that DM
\begin{recommendation}{camera_viewer}{Provide a capability for fast display of full-focal-plane images for all data ingested at the LSST Data Facility based on the LSST Camera Image Viewer}
\end{recommendation}

Users across project are most proficient with DS9. Therefore, we recommend that DM

\begin{recommendation}{ds9}{Make ds9 available through commissioning and early operations by either:
\begin{itemize}
\item{Maintaining a development server (lsst-dev equivalent) for project staff,}
\item{Connecting LSP to local ds9 via SAMP or XPA or}
\item{Through WebDAV}
\end{itemize}}
The feasibility of these options should be evaluated
\end{recommendation}

Both JS9 and AstroWidgets show promise for the Jupyterlab environment.
JS9 is snappy, has the same interface as DS9, mouse gesture stretch, and on the fly smoothing.
Ginga/AstroWidgets is snappy, extensible by python programmers, and it is easy to run arbitrary python code on selected pixels.
Therefore we recommend that DM
\begin{recommendation}{js9_ginga}{Support at least one of JS9 / AstroWidgets + Ginga on Science Platform}
A bake-off is required to determine which to support
\end{recommendation}


Finally, firefly farther ahead feature-wise.
Firefly has a few unique features.
These include its intuitive mask plane toggle and that users can open in a window on another monitor (or across the country).
Therefore we recommend that DM
\begin{recommendation}{firefly}{Continue to provide support for the Firefly tool, including some level of resourcing for servicing emergent feature requests}
\end{recommendation}


\appendix

\section{References}
\label{sec:bib}
\bibliography{lsst,lsst-dm,refs_ads,refs,books}

\section{Abbreviations and Definitions}
\label{sec:acronyms}
\input{acronyms.tex}

\end{document}
